Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-06 SCP2008/paper/metamonads.tex
partial new section layout out all the aspects in the algorithm family

@@

% previous introduction is saved at the end
\section{Introduction}

@@

The rest of this paper is structured as follows: The next section
introduces code generation in MetaOCaml, the problem of name
generation, and continuation-passing style (CPS) as a general
solution.  We also introduce the monad and the issues of generating
control statements. Section~\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
LU decomposition algorithm family in completely separate,
independent modules.  Section~\ref{s:ode} describes generating
Runge-Kutta solvers for ordinary differential equations (ODE).
We briefly discuss related work in
section~\ref{related}. We then outline the future work and conclude.
Appendices give samples of the generated code (which is available in
full at \cite{metamonadsURL}).

@@

\section{Aspects and Functors}\label{functors}
The monad represents fine-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code.

While the Object-Oriented Design community has acquired an extensive
vocabulary for describing modularity ideas, the guiding principles for
modular designs has not changed since they were first articulated by
Parnas~\cite{journals/cacm/parnas72a} and Dijkstra~
\cite{EWD:EWD447}: information hiding and separation of concerns.  To
apply these principles to the study of LU decomposition, we need
to understand what are the changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.
A study of Gaussian Elimination \cite{carette04} shows that
the following variations occur:
\begin{enumerate}
%\vspace*{-6pt}
    \item \textbf{Domain}: In which (algebraic) domain do the
      matrix elements belong to.  Sometimes the domains are very
      specific (e.g., $\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p$ and
      floating point numbers), while in other cases the domains
      were left generic, e.g., multivariate polynomials over a
      field.  In the roughly 85 pieces of code surveyed
      \cite{carette04} 20 different domains were encountered.
    \item \textbf{Container}: Whether the matrix is represented as
      an array of arrays, a one-dimensional array, a hash table, a
      sparse matrix, etc., and whether indexing is done in C or
      Fortran style.  Additionally, if a particular representation
      had a special mechanism for efficient row exchanges.
    \item \textbf{Output choices}: Whether just the reduced
      matrix, or additionally the rank, the determinant, and the
      pivoting matrix are to be returned. In the larger algorithm
      family, routines like Maple's\\
      \texttt{LinearAlgebra:-LUDecomposition} have up to $2^6 +
      2^5 + 2^2 = 100$ outputs.
    \item \textbf{Fraction-free}: Whether the Gaussian Elimination
        algorithm is allowed to use unrestricted division, or only
        exact (remainder-free) division.
    \item \textbf{Pivoting}: Whether to use no, 
        column-wise, or full pivoting.
    \item \textbf{Augmented Matrices}: Whether all or only some
      columns of the matrix participate in elimination.
\end{enumerate}
\noindent In addition to the above variations, there are two aspects that 
recur frequently:
\vspace*{-6pt}
\begin{enumerate}
    \item \textbf{Length measure}:  For stability reasons
        (numerical or coefficient growth), if a domain
      possesses
        an appropriate length measure, this is sometimes used to choose
        an ``optimal'' pivot.
    \item \textbf{Normalization and zero-equivalence}: Whether the
      arithmetic operations of the domain give normalized results, and
      whether a specialized zero-equivalence routine is to be used.
\end{enumerate}
\noindent These are separated out from the others as they are cross-cutting
concerns: in the case of the length measure, a property of the domain
will influence the pivoting method \emph{if} pivoting is to be
performed.

The simplest parametrization is to make the domain abstract. As it
turns out, we need the following to exist in our domains: $0$, $1$,
$+$, $*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.
There, code-generators such as |plus| and |one|
were passed as arguments. We need far more than
two parameters, so we have to group them. Instead of the grouping
offered by regular records, we use OCaml \emph{structures} (i.e.,
modules)
so we can take advantage of extensibility, type abstraction and constraints,
and especially parameterized structures (\emph{functors}).
We define the type of the domain, the signature |DOMAIN|, which
different domains must satisfy:

>>>>

%
% Useful abbreviations
%
\newcommand{\floats}{\mathbb{F}}
\newcommand{\reals}{\mathbb{R}}

@@

\section{Introduction}

@@

The rest of this paper is structured as follows: The next section
gives an overview of the design space of LU decomposition algorithms,
as well as the methodology we follow.  Then section~\ref{CPS}
introduces code generation in MetaOCaml, the problem of name
generation, and continuation-passing style (CPS) as a general
solution.  We also present a key monad and the issues of generating
control statements. Section~\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
LU decomposition algorithm family in completely separate,
independent modules.  Section~\ref{s:ode} describes generating
Runge-Kutta solvers for ordinary differential equations (ODE).
We briefly discuss related work in
section~\ref{related}. We then outline the future work and conclude.
Appendices give samples of the generated code (which is available in
full at \cite{metamonadsURL}).

\section{The design space}\label{design}

Before studying the details of the technologies involved in the 
implementation, it is worthwhile to carefully study the design space involved.
A preliminary study~\cite{carette04} revealed a number of aspects of the 
family of Gaussian Elimination algorithms.  In the present work, we outline
a number of additional aspects involved in the faimily of LU decomposition 
algorithms.  These will first be presented in a somewhat ad hoc manner, 
roughly coresponding to the order in which they were ``discovered''.  These
will then be reorganized into sets of semantically related aspects, which will
form the basis of our design.  Finally, we will extract a methodology from
our experience.

Throughout, we assume that the reader is familiar with the basic LU
decomposition algorithm, which factors a invertible matrix $A$ into a unit
lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
such that $A = LU$.  When pivoting is used, we get a unitary matrix $P$ such
that the factorization is now $A = PLU$.  The case of numeric matrices is well
covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
a $PLU$ decomposition, with $L$ remaining unit lower-triangular but with a
$U$ which is no longer upper triangular but instead ``staggered''
in the upper triangle.

\subsection{Aspects}

We reuse the english word ``aspect'' for the various facets of the family
of LU decomposition algorithms.  While this use if very much in the same 
spirit as in aspect-oriented programming (AOP)~\cite{Kiczales}, our 
implementation methodology is radically different.  We firmly believe that
our typed generative methodogy is much better suited to the functional
programming paradigm than attempting to graft the program-trace-based
methodology of object-oriented versions of AOP.

At this point in time, it is better to think of aspects as purely
design-time entities.  Here we are firmly influenced by Parnas' original
view of modules and information hiding~\cite{journals/cacm/parnas72a} as well
as his view of product families \cite{Parnas??}, and by Dijkstra's ideas on
separation of concerns \cite{EWD:EWD447}.
To apply these principles to the study of LU decomposition, we need
to understand what are the changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.

The various aspects listed below all come from variations found in actual
implementations (in various languages and settings).

\begin{enumerate}
	\item \textbf{Domain}: In which (algebraic) domain do the matrix
		elements belong to.  Some implementations were very specific
		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z_p}, 
		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
		\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$, 
		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers 
		($\floats$) for 
		example), while others were generic for elements of a field,
		multivariate polynomials over a field, or elements of a division ring
		with possibly undecidable zero-equivalence.  In the roughly 85 pieces
		of code we surveyed, 20 different domains were encountered.
	\item \textbf{Representation of the matrix}: Whether the matrix
		was represented as an array of arrays, a one-dimensional array,
		a hash table, etc.  For the case of a one-dimensional array,
		whether indexing was done in C or Fortran style.  Additionally,
		if a particular representation had a special mechanism for efficient
		row exchanges, this was sometimes used.
	\item \textbf{Fraction-free}: Whether the 
		algorithm is allowed to use unrestricted division, or only
		exact (remainder-free) division.
	\item \textbf{Length measure (for pivoting)}:  For stability reasons
		(whether numerical or coefficient growth), if a domain possesses
		an appropriate length measure, this was sometimes used to choose
		an ``optimal'' pivot.  Not all domains have such a measure.
	\item \textbf{Full division}: Whether the input domain supports full
		division (i.e. is a \emph{field} or pretends to be ($\floats$)) 
		or only exact division (i.e. a \emph{division ring}).
	\item \textbf{Domain normalization}: Whether the arithmetic operations
		of the base domain keep the results in normal form, or whether
		an extra normalization step is required.  For example, some 
		representations of polynomials require an extra step for
		zero-testing.
	\item \textbf{Output choices}:  Whether just the reduced matrix
		(the `U' factor), both L and U, as well as
		the rank, the determinant, and/or the sequence of 
		pivots is to be returned.  For example, Maple's
		\texttt{LinearAlgebra:-LUDecomposition} routine has
		$2^6 + 2^5 + 2^2 = 100$ possible outputs, depending on whether
		one chooses a $PLU$, $PLU^{1}R$ or \emph{Cholesky} 
		decomposition.  We chose to only consider $PLU$ for now.
	\item \textbf{Rank}: Whether to explicitly track the rank of the matrix
		as the algorithm proceeds.
	\item \textbf{Determinant}:  Whether to explicitly track the determinant
		of the matrix as the algorithm proceeds.
	\item \textbf{Code representation}: Whether we are actually generating
		code or we are in fact representation the algorithm in ``direct
		style'', albeit with a lot of abstraction overhead.
	\item \textbf{Zero-equivalence}: Whether the 
		arithmetic operations require a specialized zero-equivalence 
		routine needs to be used.  It turns out that for certain classes
		of \emph{expressions}, it is convenient to use a zero-equivalence
		test which is separate from the domain normalization.  This is
		usually the case when zero-equivalence is formally undecidable
		but semi-algorithms or probabilistic algorithms do exist.
		See~\cite{ZhCaJeMo07} for an example.
    \item \textbf{Pivoting}: Whether to use no, 
        column-wise, or full pivoting.
    \item \textbf{Augmented Matrices}: Whether all or only some
      columns of the matrix participate in elimination.
	\item \textbf{Pivot representation}: Whether the pivot is represented
	  as a list of row (and/or column) exchanges, as a unitary matrix,
	  or as a permutation vector.
  \item \textbf{Lower matrix}: whether the matrix $L$ should be tracked
	  as the algorithm proceeds, reconstructed at the end of the
	  algorithm, or not tracked at all.
  \item \textbf{Input choices}: Whether the input matrix is an augmented
	  matrix or not, and if so, where is the end of the main matrix.
  \item \textbf{Packed}: Whether the output matrices $L$ and $U$ are
	  packed into a single matrix for output.
\end{enumerate}

\noindent \jacques{Absorb this in the above} These are separated out from the
others as they are cross-cutting concerns: in the case of the length measure, a
property of the domain will influence the pivoting method \emph{if} pivoting is
to be performed.\jacques{Also, should mention a few more aspects like
inplace, errors-on-singular, logging, sparsity, as things that are not
implemented}

\subsection{Organizing aspects}

\subsection{Methodology}

@@

\section{Aspects and Functors}\label{functors}

The monad represents fine-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code.

Reviewing the various aspects outlined in section~\ref{aspects},
the simplest parametrization is to make the domain abstract. 
\jacques{This whole section needs to be edited to take into account
the different organization of the material}.
As it
turns out, we need the following to exist in our domains: $0$, $1$,
$+$, $*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. The simplest
case of such domain abstraction is |param_code1|.
There, code-generators such as |plus| and |one|
were passed as arguments. We need far more than
two parameters, so we have to group them. Instead of the grouping
offered by regular records, we use OCaml \emph{structures} (i.e.,
modules)
so we can take advantage of extensibility, type abstraction and constraints,
and especially parameterized structures (\emph{functors}).
We define the type of the domain, the signature |DOMAIN|, which
different domains must satisfy:
